An investigation into the contentious phylogenetic results of the *Mixta* species

### 2021-05-02 Manuscript/Manuscript_Paper.Rmd
Update Manuscript_Paper.Rmd

<!-- ## Nucleotide and Amino Acid Identity -->
<!-- Each of the non-*Mixta* sequences were aligned to one of the corresponding *Mixta* sequences in turn. Alignments were performed in R as described above, and the aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between each sequence and the *Mixta* homologs. Values that equal or are close to 100% are more identical to the *Mixta* sequences. -->

@@

## Closest Relative from Average Site Identity
Site identity is the amount of similarity between two sequences and accounts for the sequence length. The non-*Mixta* sequences were aligned to each of the *Mixta* homologs, so the percent identity was relative to *Mixta*. The closest relative to both *Mixta* species is defined as the taxon with the highest similarity to *Mixta*.

@@

## Spatial Patterns from Average Site Identity
Positional effects were investigated in the site identity results. Only the results of *M. calida* are shown because the *M. gaviniae* results showed similar patterns. The genes most similar to either of the *Pantoea* species and *E. cloacae* were evenly dispersed across the *Mixta* chromosome for both the NTS and AAS (Figure 5A and 5B). Similar to the genetic distance results, genes that are labelled as being closely related to either of the *Tatumella* species are located only on right side of chromosome, and genes that are closely related to *E. amylovora* are predominantly on the right side as well (Figure 5A). Unexpectedly, the lower left quadrant of the chromosome contains primarily genes that are closely related to either of the *Pantoea* species with only a few exceptions (Figure 5A). 	In the AAS analysis, genes closely related to any of the other taxa were more evenly distributed across the genome (Figure 5B).

@@

## A Comparison of the Four Analyses
The four above analyses (genetic distance and site identity, both using NTS and AAS) gave slightly different results for the *Mixta* sequences. The four analyses agreed upon a closest relative for approximately 38.7% of each of the *M. calida* and *M. gaviniae* sequences. The number of sequences that were closely related to one of the other taxa differed between the two *Mixta* species (Table 3). Additionally, although 38.7% of both *Mixta* species’ sequences had a consistent closest relative, 30 genes differed between them. For these 30 genes, there was a consensus for only one of the *Mixta* species and not the other. Nevertheless, 34.9% of the *M. calida* and *M. gaviniae* homologs agreed on the same species.

@@

Displaying these results in a circular plot showed that genes closely related to any of the taxa were dispersed across the *M. calida* and *M. gaviniae* chromosomes (Figures 6A and 6B, respectively). Given that similar patterns were found in the four analyses, genes with a consistent closest relative are randomly distributed across the genome and the position within the genome does not have an effect on whether the analyses agree (Figures 6A and 6B).

@@

my_colours <- c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", 
                "#E31A1C", "#FDBF6F") # For P_septica, P_agglomerans, E_tasmaniensis, E_amylovora, T_saanichensis, and E_cloacae

@@

Interestingly, the four analyses agreed on a genus being the closest relative for a total of 63.4 and 63.2% of the *M. calida* and *M. gaviniae* sequences, respectively (Table 4). Unexpectedly, only 39 of the *M. calida* and the *M. gaviniae* homologs did not agree on which genus was the closest relative. Of these genes, there was a consensus for one of *M. calida* or *M. gaviniae* and not the other for 38 homologs. For only one gene (*ttuB*) did the analyses agree on *Tatumella* for *M. calida* and *Pantoea* for *M. gaviniae.

@@

## Phylogenetic Trees

>>>>

<!-- ## Nucleotide and Amino Acid Identity -->
<!-- Each of the non-*Mixta* sequences were aligned to one of the corresponding *Mixta* sequences in turn. Alignments were performed in R as described above, and the aligned sequences were then split into the individual sites using the `str_split()` function (package: stringr) [@Wickham2019]. A function was created using base `R` code to calculate the number of identical sites between each sequence and the *Mixta* homologues. Values that equal or are close to 100% are more identical to the *Mixta* sequences. -->

@@

## Closest Relative from Average Site Identity
Site identity is the amount of similarity between two sequences and accounts for the sequence length. The non-*Mixta* sequences were aligned to each of the *Mixta* homologues, so the percent identity was relative to *Mixta*. The closest relative to both *Mixta* species is defined as the taxon with the highest similarity to *Mixta*.

@@

## Spatial Patterns from Average Site Identity
Positional effects were investigated in the site identity results. Only the results of *M. calida* are shown because the *M. gaviniae* results showed similar patterns. The genes most similar to either of the *Pantoea* species and *E. cloacae* were evenly dispersed across the *Mixta* chromosome for both the NTS and AAS (Figure 5A and 5B). Similar to the genetic distance results, genes that are labelled as closely related to either of the *Tatumella* species are located only on the right side of the chromosome, and genes closely related to *E. amylovora* are predominantly on the right side as well (Figure 5A). Unexpectedly, the lower left quadrant of the chromosome primarily contains genes closely related to either of the *Pantoea* species, with only a few exceptions (Figure 5A). In the AAS analysis, genes closely related to any of the other taxa were more evenly distributed across the genome (Figure 5B).

@@

## A Comparison of the Four Analyses
The four above analyses (genetic distance and site identity, both using NTS and AAS) gave slightly different results for the *Mixta* sequences. The four analyses agreed upon the closest relative for approximately 38.7% of each of the *M. calida* and *M. gaviniae* sequences. The number of sequences closely related to one of the other taxa differed between the two *Mixta* species (Table 3). Additionally, although 38.7% of both *Mixta* species’ sequences had a consistent closest relative, 30 genes differed between them. For these 30 genes, there was a consensus for only one of the *Mixta* species and not the other. Nevertheless, 34.9% of the *M. calida* and *M. gaviniae* homologues agreed on the same species.

@@

Displaying these results in a circular plot showed that genes closely related to any of the taxa were dispersed across the *M. calida* and *M. gaviniae* chromosomes (Figures 6A and 6B, respectively). Given that similar patterns were found in the four analyses, genes with a consistent closest relative are randomly distributed across the genome. The position within the genome does not affect whether the analyses agree (Figures 6A and 6B).

@@

my_colours <- c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", # For P_septica, P_agglomerans, E_tasmaniensis, E_amylovora,
                "#E31A1C", "#FDBF6F") # T_saanichensis, and E_cloacae

@@

Interestingly, the four analyses agreed on a genus being the closest relative for a total of 63.4 and 63.2% of the *M. calida* and *M. gaviniae* sequences, respectively (Table 4). Unexpectedly, only 39 of the *M. calida* and the *M. gaviniae* homologues did not agree on which genus was the closest relative. Of these genes, there was a consensus for one of *M. calida* or *M. gaviniae* and not the other for 38 homologues. Only one gene (*ttuB*) did the analyses agree on *Tatumella* for *M. calida* and *Pantoea* for *M. gaviniae.

@@

## Phylogenetic Gene Trees for *secY* and *ttuB*
In the genetic distance analysis, only one (*ttuB*) and two (*secY* and *ttuB*) genes differed between the two *Mixta* species for the NTS and AAS, respectively. The phylogenetic trees for the *secY* gene show that the gene is well-conserved across the *Erwiniaceae* taxa (Figures 7A and 7B). When using the NTS, the *Mixta* species are grouped together and fall outside of the other three *Erwiniaceae* taxa (Figure 7A). When using the AAS, the *Mixta* species do not group together (Figure 7B). Additionally, *M. calida* has a slightly shorter branch length to the *Erwinia* species, whereas *M. gaviniae* has a somewhat shorter branch length to *P. septica* (Figure 7B). However, the nodes separating the *Mixta* species from the other taxa are not well-supported (Figures 7A and 7B). Unexpectedly, *P. septica*’s homologue of *secY* is most closely related to *E. cloacae* for the AAS (Figure 7B) whereas *E. cloacae* falls outside of the *Erwiniaceae* for the NTS (Figure 7A).

![Figure 7. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringiae* as outgroups. The trees were reconstructed from the nucleotide (**A**) and amino acid (**B**) sequences of the gene *secY*. HKY+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure7_secY_Trees.png)

The phylogenetic trees for the *ttuB* gene show a clearer separation between the two *Mixta* homologues (Figures 8A and 8B). For both the NTS and AAS, *M. calida* was more closely related to *Tatumella* followed by *E. cloacae* and *M. gaviniae* was more closely related to *Pantoea* followed by *Erwinia* (Figures 8A and 8B). Unlike with *secY*, many of the nodes are well-supported (Figures 8A and 8B).

![Figure 8. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringiae* as outgroups. The trees were reconstructed from the nucleotide (**A**) and amino acid (**B**) sequences of the gene *ttuB*. GTR+G and LG+G models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure8_ttuB_Tree.png)

## Phylogenetic Trees from MLSA Genes
Phylogenetic analyses often use several concatenated essential gene sequences to infer relatedness, namely *atpD*, *gyrB*, *infB*, and *rpoB*. However, *atpD* and *gyrB* did not meet the length requirements during the filtering process. Therefore, genes encoding the other subunits of ATP synthase (*atp*), translation initiation factor IF (*inf*), and DNA-directed RNA polymerase (*rpo*) were concatenated to create phylogenetic trees. The resulting trees show that *Mixta* are more closely related to *Pantoea* followed by *Erwinia* (Figures 9A and 9B). With the NTS, a well-supported bifurcation occurs between *Mixta* and *Pantoea* (Figure 9A). However, this does not occur with the AAS (Figure 9B). Instead, a somewhat-supported bifurcation occurs first between *Pantoea* and *Erwinia* and then a little-supported bifurcation between these two and *Mixta* (Figure 9B).

![Figure 9. Maximum-likelihood phylogenetic trees for members of the *Erwiniaceae* with *E. cloacae* and *P. syringiae* as outgroups. The trees were reconstructed from the concantenated nucleotide (**A**) and amino acid (**B**) sequences of the genes *atpA*, *atpB*, *atpC*, *atpE*, *atpF*, *atpG*, *atpH*, *infA*, *infB*, *rpoA*, *rpoB*, *rpoC*, *rpoD*, and *rpoH*. GTR+G+I and LG+G+F models were used for the nucleotide and amino acid sequences, respectively. Both trees required a discrete Gamma distribution (5 categories) to model the evolutionary rate differences among sites. The nucleotide model allowed for some sites to be evolutionary invariable. Branch support was inferred from 500 bootstrap replicates; the percentage of trees in which the associated taxa clustered together are adjacent to the branches. The tree is drawn to scale, with branch lengths measured in the number of substitutions per site. Evolutionary analyses were conducted in MEGAX.](C:/Users/Kim/OneDrive/2020_3Fall/Biology_396/10_Tables_and_Figures/Figure9_Four_Tree.png)

The four analyses conducted in this study agreed on a closest relative for only four of these genes – *atpA*, atpG*, *rpoB*, and *rpoC* – for both *M. calida* and *M. gaviniae*. This closest relative was *P. septica* for these genes. The other genes were more closely related to *P. agglomerans*, *E. tasmaniensis*, *E. amylovora*, and *E. cloacae* and these varied amongst the analyses. The gene *rpoD* was the only gene that was said to be closely related to *E. cloacae* by one of the analyses. The NTS genetic distance analysis labelled both the *Mixta* species as being closely related to *E. cloacae* and the remaining three analyses labelled this gene as being closely related to *P. septica*.

## Phylogenetic Trees Showing Other Results
